Measuring matter effects in Neutron star - Black hole binaries with Advanced LIGO

### 2016-04-07 Paper/paper.tex
minor changes

Coalescing binaries of compact objects, i.e. black holes (BH) and/or neutron 
stars (NS), remain the most interestings targets for GW based observation with
aLIGO. NSBH binaries form an important sub-class of compact binaries, and we
expect to observe $\mathcal{O}(10)$ of these every year with design 
sensitivity aLIGO~\cite{Abadie:2010cf}.
% 
NSBH mergers are of interest for various reasons. For instance, they have 
been associated with (as the progenitors of) short Gamma-ray Bursts
(GRBs)~\cite{eichler:89,1992ApJ...395L..83N,moch:93,Barthelmy:2005bx,
2005Natur.437..845F,2005Natur.437..851G,Shibata:2005mz,Paschalidis2014,
Tanvir:2013}. Under favorable conditions, close to merger the neutron star
can get disrupted by the tidal field of its companion hole. Once disrupted,
the neutron star material may form an accretion disk around the black hole. 
Most of the disk matter spirals in within a few milliseconds, with a small 
fraction becoming unbounded and getting ejected, with a possibled simultaneous
burst of gamma-ray emission)~\cite{Foucart2015,Lovelace:2013vma,Deaton2013,
Foucart2012,Shibata:2005mz,Paschalidis2014}. A coincident detection of GWs
from an NSBH system with a GRB would therefore provide us insights into the
central engine powering GRBs. 
% 
Another question that NSBH mergers can help answer is `what is the 
nature of matter at nuclear densities', i.e. what is the equation of state
(EoS) of neutron stars. This is what we will focus on in this paper.
% 
The gravitational-waves emitted by NSBH binaries carry subtle hints of the 
matter structure of the neutron star. During early inspiral, the (weak) tidal
field of the companion hole deforms the star, thereby exciting its resonant
oscillation modes. This drains energy from the orbit resulting in an
increase of the inspiral rate. We know the leading and next-to-leading
order terms in post-Newtonian (PN) theory that capture this 
effect~\cite{Vines2011}, entering binary phasing at $5$PN order. Close to
merger, as the NS gets disrupted and forms long tails that feed into the 
companion black hole, the quadrupole moment of the system falls monotonically.
As a result, the emitted GWs are subdued significantly post-disruption
(as compared to what they would be for a BBH),
and the quasi-normal modes of the resulting black hole are not 
excited~\cite{Kyutoku:2010zd,Lackey:2013axa,Lovelace:2013vma,Foucart2015,
Pannarale:2015jia}.
% 
% 
The difficulty in discerning NS matter effects in an NSBH coalescence from its
GW signal lies in distinguishing it reliably from a BBH signal, since the 
difference between the two is either subtle (during inspiral) or short-lived
(disruption and merger). The earlier in inspiral the NS disrupts, the more
orbits there are over which its effect is visible. The mechanism of disruption
is governed by the nature of the BH, i.e. its mass and spin. More massive BHs 
(with $m_\mathrm{BH}\gtrsim 7\, m_\mathrm{NS}$), as well as BHs with
(large) retrograde spins $(\chi_\mathrm{BH}\in[-1. 0])$, tend to swallow the 
NS whole due to a rapid merger (i.e. without it disrupting before the
inner-most stable circular orbit (ISCO) is reached). The quasi-normal
oscillation modes of the final black hole's horizon are also 
excited~\cite{Foucart:2013psa}. Therefore, the only available information 
comes from the inspiral rate modification. On the other hand, less massive
black holes spinning in the prograde sense (with mass-ratio 
$q\in[2, 5]$, and $\chi_\mathrm{BH}\geq 0$) can cause their companion star to 
disrupt much before merger, leaving a strong imprint on the emitted GW 
signal~\cite{Shibata:2007zm,2010PhRvD..81f4026F,Lovelace:2013vma,
Kawaguchi:2015}.

@@

As this line of inquiry relies mainly on 
classifying a GW signal as an NSBH binary, it can be pursued using non-tidal
waveform models~\cite{}.

% Conventional astronomical methods have, to date, made observations of ~2500
% NSs with masses between $1.25-2.1 M_\odot$~\cite{Lyne:2004cj,
% Demorest:2010bx,2013Sci...340..448A,atnfcatalog,mcgillmagnetarcatalog,
% stellarcollapsemass}, although
% tightly clustered within $1.35\pm0.5M_\odot$~\cite{stellarcollapsemass}.
% %
% Observations of over 2500 NSs have shown that the spin of NSs
% $|\vec{\chi}_\mathrm{NS}|$ have magnitudes that are $<1\%$~\cite{Miller:2014aaa}.
% %
% On the other hand, indirect observations of stellar-mass BHs place their
% masses between $5-35M_\odot$, with their spin angular momenta 
% $|\vec{\chi}_\mathrm{BH}|$ ranging from small to nearly extremal (Kerr) values
% (see, e.g., Refs.~\cite{McClintockEtAl:2006,Miller:2009cw,Gou:2014una} for 
% examples of nearly extremal estimates of BH spins, Refs.~\cite{McClintock:2013vwa,
% Reynolds:2013qqa} for recent reviews of astrophysical BH spin measurements,
% and Figure 5 of Ref.~\cite{Miller:2014aaa} for a comparison of NS and BH spins).
%

Finally, we note that our results are applicable for the design sensitivity
LIGO instruments, that are expected to come online in $2018-19$. Also note
that more a recent work that improves upon the waveform model we use in this
study has been published~\cite{Pannarale:2015jka}, but it gives only an 
amplitude model and needs to be (in future) augmented with a compatible phase
model. In addition, accuracy of the underlying numerical simulations used to
calibrate the waveform model used here has not been verified against 
independent codes so far.
It is therefore difficult to assess the combined modeling error and the 
results presented in our study are limited by the limitations of the model.
However, we expect the combined effect of modeling errors to not change our
qualitative conclusions.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}\label{s1:discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Discussion

>>>>

Coalescing binaries of compact objects, i.e. black holes (BH) and/or neutron 
stars (NS), remain the most interestings targets for GW based observation with
aLIGO. NSBH binaries form an important sub-class of compact binaries, and we
expect to observe $\mathcal{O}(10)$ of these every year with design 
sensitivity aLIGO~\cite{Abadie:2010cf}.
NSBH mergers are of interest for various reasons. For instance, they have 
been associated with (as the progenitors of) short Gamma-ray Bursts
(GRBs)~\cite{eichler:89,1992ApJ...395L..83N,moch:93,Barthelmy:2005bx,
2005Natur.437..845F,2005Natur.437..851G,Shibata:2005mz,Paschalidis2014,
Tanvir:2013}. Under favorable conditions, close to merger the neutron star
can get disrupted by the tidal field of its companion hole. Once disrupted,
the neutron star material may form an accretion disk around the black hole. 
Most of the disk matter spirals in within a few milliseconds, with a small 
fraction becoming unbounded and getting ejected, with a possibled simultaneous
burst of gamma-ray emission)~\cite{Foucart2015,Lovelace:2013vma,Deaton2013,
Foucart2012,Shibata:2005mz,Paschalidis2014}. A coincident detection of GWs
from an NSBH system with a GRB would therefore provide us insights into the
central engine powering GRBs. 
Another question that NSBH mergers can help answer is `what is the 
nature of matter at nuclear densities', i.e. what is the equation of state
(EoS) of neutron stars. This is what we will focus on in this paper.
The gravitational-waves emitted by NSBH binaries carry subtle hints of the 
matter structure of the neutron star. During early inspiral, the (weak) tidal
field of the companion hole deforms the star, thereby exciting its resonant
oscillation modes. The induced quadrupolar moment of the star depends on the
tidal field $\mathcal{E}_{ij}$ due to its companion through the relation
$Q_{ij} = -\lambda\mathcal{E}_{ij}$, where $\lambda$ is the EoS dependent
tidal deformability parameter related to the neutron star's dimensionless
Love number $k_2$ and radius $R_\mathrm{NS}$ as
$\lambda:=\frac{2}{3G}k_2 R_\mathrm{NS}^5$. This drains energy from the orbit
resulting in an increase of the inspiral rate. We know the leading and next-to
-leading order terms in post-Newtonian (PN) theory that capture this 
effect~\cite{Vines2011}, entering binary phasing at $5$PN order. Close to
merger, as the NS gets disrupted and forms long tails that feed into the 
companion black hole, the quadrupole moment of the system falls monotonically.
As a result, the emitted GWs are subdued significantly post-disruption
(as compared to what they would be for a BBH),
and the quasi-normal modes of the resulting black hole are not 
excited~\cite{Kyutoku:2010zd,Lackey:2013axa,Lovelace:2013vma,Foucart2015,
Pannarale:2015jia}.
The difficulty in discerning NS matter effects in an NSBH coalescence from its
GW signal lies in distinguishing it reliably from a BBH signal, since the 
difference between the two is either subtle (during inspiral) or short-lived
(disruption and merger). The earlier in inspiral the NS disrupts, the more
orbits there are over which its effect is visible. The mechanism of disruption
is governed by the nature of the BH, i.e. its mass and spin. More massive BHs 
(with $m_\mathrm{BH}\gtrsim 7\, m_\mathrm{NS}$), as well as BHs with
(large) retrograde spins $(\chi_\mathrm{BH}\in[-1. 0])$, tend to swallow the 
NS whole due to a rapid merger (i.e. without it disrupting before the
inner-most stable circular orbit (ISCO) is reached). The quasi-normal
oscillation modes of the final black hole's horizon are also 
excited~\cite{Foucart:2013psa}. Therefore, the only available information 
comes from the inspiral rate modification. On the other hand, less massive
black holes spinning in the prograde sense (with mass-ratio 
$q\in[2, 5]$, and $\chi_\mathrm{BH}\geq 0$) can cause their companion star to 
disrupt much before merger, leaving a strong imprint on the emitted GW 
signal~\cite{Shibata:2007zm,2010PhRvD..81f4026F,Lovelace:2013vma,
Kawaguchi:2015}.

@@

Finally, we note that our results are applicable for the design sensitivity
LIGO instruments, that are expected to come online in $2018-19$. Also note
that more a recent work that improves upon the waveform model we use in this
study has been published~\cite{Pannarale:2015jka}, but it provides only an 
amplitude model which needs to be (in future) augmented with a compatible
phase model. In addition, accuracy of the underlying numerical simulations 
used to calibrate the waveform model used here have not been verified against 
independent codes so far.
It is therefore difficult to assess the combined modeling error and its effect
on our results. they are, therefore, limited by the limitations of our
waveform model. However, we expect the combined effect of modeling errors to
not change our qualitative conclusions.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}\label{s1:discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Conventional astronomical methods have, to date, made observations of ~2500
NSs with masses between $1.25-2.1 M_\odot$~\cite{Lyne:2004cj,
Demorest:2010bx,2013Sci...340..448A,atnfcatalog,mcgillmagnetarcatalog,
stellarcollapsemass}, although
tightly clustered within $1.35\pm0.5M_\odot$~\cite{stellarcollapsemass}.
%
Observations of over 2500 NSs have shown that the spin of NSs
$|\vec{\chi}_\mathrm{NS}|$ have magnitudes that are $<1\%$~\cite{Miller:2014aaa}.
%
On the other hand, indirect observations of stellar-mass BHs place their
masses between $5-35M_\odot$, with their spin angular momenta 
$|\vec{\chi}_\mathrm{BH}|$ ranging from small to nearly extremal (Kerr) values
(see, e.g., Refs.~\cite{McClintockEtAl:2006,Miller:2009cw,Gou:2014una} for 
examples of nearly extremal estimates of BH spins, Refs.~\cite{McClintock:2013vwa,
Reynolds:2013qqa} for recent reviews of astrophysical BH spin measurements,
and Figure 5 of Ref.~\cite{Miller:2014aaa} for a comparison of NS and BH spins).
